\pkg{depmixS4}: An \proglang{R}-package for hidden Markov models

### 2009-07-13 papers/jss/article.tex
Added all the code examples, albeit with minimal text.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@293 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% an abstract and keywords
\Abstract{
	\pkg{depmixS4} implements a general framework for definining and
	fitting hidden Markov mixture model in the R programming language
	\citep{R2009}.  This includes standard Markov models,
	latent/hidden Markov models, and latent class and finite mixture
	distribution models.  The models can be fitted on mixed
	multivariate data with multinomial and/or gaussian distributions.
	Parameters can be estimated subject to general linear constraints.
	Parameter estimation is done through an EM algorithm or by a
	direct optimization approach using the
	\pkg{Rdonlp2} optimization routine when contraints are imposed on
	the parameters.  A number of illustrative examples are included.

}
\Keywords{hidden Markov model, dependent mixture model, mixture model, \proglang{R}}
\Plainkeywords{hidden Markov model, dependent mixture model, mixture model,R} %% without formatting
%% at least one keyword must be supplied

@@

\begin{figure}[htbp]
  \begin{center}
	  \scalebox{1}{\includegraphics*{graphs/speed1.pdf}}
	  \caption{Reaction times, accuracy and pay-off values for
	  the first series of responses in dataset \texttt{speed}.}
	  \label{fig:speed}
  \end{center}
\end{figure}

@@

Need model equations: maybe from Fruhwirth-Schnatter?

@@

Throughout this manual a data set called \code{speed} is used.  It
consists of three time series with three variables: reaction time,
accuracy, and a covariate Pacc which defines the relative pay-off for
speeded and accurate responding. The participant in this experiment 
switches between fast responding at chance level and relatively 
slower responding at a high level of accuracy.

Interesting hypotheses to test are: is the switching regime symmetric?
Is there evidence for two states or does one state suffice?  Is the
guessing state actually a guessing state, i.e., is the probability
of a correct response at chance level (0.5)?

@@

A dependent mixture model is defined by the number of states, and by
the response distribution functions, and can be created with the
\code{depmix}-function as follows (see the help files for other
options):

\begin{verbatim} 
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, trstart=runif(4))
fm <- fit(mod)
\end{verbatim}

The function returns an object of class \code{depmix.fitted} which
extends the \code{depmix} class, adding convergence information (and
information about constraints if these were applied, see section on Constraints below).  The
\code{print} method provides summary information on convergence, the log likelihood
and the AIC and BIC values. These statistics may be extracted using  \code{logLik},
\code{AIC} and \code{BIC}, respectively.

@@

The \code{summary} method of fitted models provides the parameter estimates, first for
the prior probabilities model, second for the transition model, and third for the response models.

@@

\subsection{Adding covariates on transition parameters}

The transition matrix is parametrized as a list of multinomial
logistic models.  The initial state probabilities are similarly
parametrized as a multinomial logistic model.  Both models use a baseline
category parametrization, meaning that the parameter for the base
category is fixed at zero.  The default base category is the first
state.  Hence, for example, for a 3-state model, the initial state
probability model would have three parameters of which the first is
fixed at zero and the other two are freely estimated.

See \citet[see][p.\ 267 ff.]{Agresti2002} for multinomial logistic models and various
parameterizations.

Covariates can be specified using a one-sided formula as in the
following example:

\begin{verbatim}
set.seed(1)
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, family=list(gaussian(), multinomial()),
transition=~Pacc, instart=runif(2))
fm <- fit(mod)
\end{verbatim}

\subsection{Add covariates to prior model}

Add example of balance data.

Note that this can be done for the initial state probabilities by
specifying \code{prior=~X1}, where X1 is the desired covariate.  The result
of this is that the transition probabilities are now dependent on the
covariate Pacc (which is an experimenter controlled variable to induce
switching between guessing and accurate responding).

\begin{verbatim}
data(balance)
set.seed(1)
mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
        family=list(multinomial(),multinomial(),multinomial(),multinomial()),
        respstart=runif(16))
fm <- fit(mod)
\end{verbatim}

@@

Using package \pkg{Rdonlp2}, parameters may be fitted subject to 
general linear (in-)equality constraints. In particular, the following
constraints may be applied:

Constraining and fixing parameters is done using the \code{conpat}
argument to the \code{depmix.fit}-function, which specifies for each
parameter in the model whether it's fixed (0) or free (1 or higher).
Equality constraints can be imposed by having two parameters have the
same number in the \code{conpat} vector.  When only fixed values are
required the \code{fixed} argument can be used instead of
\code{conpat}, with zeroes for fixed parameters and other values (ones
e.g.) for non-fixed parameters.  Fitting the models subject to these
constraints is handled by the optimization routine \code{donlp2}.

\paragraph{Parameter numbering} When using the \code{conpat} and
\code{fixed} arguments, complete parameter vectors should be supplied,
i.e., these vectors should have length of the number of parameters of
the model, which can be obtained by calling \code{npar(object)}.
Parameters are numbered in the following order:
\begin{enumerate}
	\item  the prior model parameters

	\item  the parameters for the transition models

	\item  the response model parameters per state (and subsequently
	per response in the case of multivariate time series)

To see the ordering of parameters use the following:
\begin{verbatim}
mod <- setpars(mod, value=1:npar(mod))mod
\end{verbatim}

To see which parameters are fixed (by default only baseline parameters
in the multinomial logistic models for the transition models and the
initial state probabilities model:
\begin{verbatim}
mod <- setpars(mod,
getpars(mod,which="fixed"))mod
\end{verbatim}

@@

Introduce the exgaus example.

Explain the use of makeDepmix for having full control of every aspect of the model.

>>>>

%% an abstract and keywords
	\Abstract{

	\pkg{depmixS4} implements a general framework for definining and
	fitting hidden Markov mixture model in the R programming language
	\citep{R2009}.  This includes standard Markov models,
	latent/hidden Markov models, and latent class and finite mixture
	distribution models.  The models can be fitted on mixed
	multivariate data with multinomial and/or gaussian distributions.
	Parameters can be estimated subject to general linear constraints.
	Parameter estimation is done through an EM algorithm or by a
	direct optimization approach using the \pkg{Rdonlp2} optimization
	routine when contraints are imposed on the parameters.  A number
	of illustrative examples are included.

}

\Keywords{hidden Markov model, dependent mixture model, mixture model, \proglang{R}}

\Plainkeywords{hidden Markov model, dependent mixture model, mixture model,R} %% without formatting

%% at least one keyword must be supplied

@@

\begin{figure}[htbp]
  \begin{center}
	  \scalebox{1.2}{\includegraphics*{graphs/speed1.pdf}}
	  \caption{Response times, accuracy and pay-off values for
	  the first series of responses in dataset \texttt{speed}.}
	  \label{fig:speed}
  \end{center}
\end{figure}

@@

The dependent mixture model is then given by the following equations: 
\begin{align}
	S_{t} = A S_{t-1}, t=2, \ldots, T \\
	O_{t} = f(O_{t}|S_{t}),
\end{align}
where $S_{t}$ is a sequence of hidden states, $A$ is a transition
matrix, $O_{t}$ is an (possibly multivariate) observation and $f$ is a
density function for $O_{t}$ conditional on the hidden state $S_{t}$.
In the example data above, $f$ could be a gaussian distribution
function for the response time variable, and a Bernouilli for the
accuracy data.  In the models we are considering here, both the
transition probabilities $A$ and the initial state probabilities $\pi$
may depend on covariates as well as the response distributions $f$. 
See for example \citet{Fruhwirth-Schnatter2006} for an overview of 
hidden Markov models with extensions.

@@

Throughout this manual a data set called \code{speed} is used.  It
consists of three time series with three variables: response time,
accuracy, and a covariate Pacc which defines the relative pay-off for
speeded and accurate responding.  The participant in this experiment
switches between fast responding at chance level and relatively slower
responding at a high level of accuracy.  Interesting hypotheses to
test are: is the switching regime symmetric?  Is there evidence for
two states or does one state suffice?  Is the guessing state actually
a guessing state, i.e., is the probability of a correct response at
chance level (0.5)?

@@

A dependent mixture model is defined by the number of states, and by
the response distribution functions, and can be created with the
\code{depmix}-function as follows:

\begin{verbatim} 
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, trstart=runif(4))
\end{verbatim}
The \code{depmix} function returns an object of class \code{depmix}
which contains the model specification (and not a fitted model!).
Note also that start values for the transition parameters are provided
in this call using the \code{trstart} argument. The package does not 
provide automatic starting values.

The so-defined models needs to be \code{fit}ted with the following: 
\begin{verbatim}
fm <- fit(mod)
\end{verbatim}

The \code{fit}-function returns an object of class
\code{depmix.fitted} which extends the \code{depmix} class, adding
convergence information (and information about constraints if these
were applied, see below).  The \code{print} method provides summary
information on convergence, the log likelihood and the AIC and BIC
values.  These statistics may be extracted using \code{logLik},
\code{AIC} and \code{BIC}, respectively.

@@

The \code{summary} method of \code{fit}ted models provides the parameter
estimates, first for the prior probabilities model, second for the
transition model, and third for the response models.

@@

\subsection{Transition parameters}

The transition matrix is parametrized as a list of multinomial
logistic models.  The initial state probabilities are similarly
parametrized as a multinomial logistic model.  Both models use a
baseline category parametrization, meaning that the parameter for the
base category is fixed at zero (see \citet[see][p.\ 267
ff.]{Agresti2002} for multinomial logistic models and various
parameterizations).  The default base category is the first state.
Hence, for example, for a 3-state model, the initial state probability
model would have three parameters of which the first is fixed at zero
and the other two are freely estimated.

Covariates on the transition probabilities can be specified using a
one-sided formula as in the following example:
\begin{verbatim}
set.seed(1)
mod <- depmix(rt~1, data=speed, nstates=2, 
family=gaussian(), transition=~scale(Pacc), 
instart=runif(2))
fm <- fit(mod)
\end{verbatim}

\begin{verbatim}
	Initial state probabilties model 
	Model of type multinomial, formula: ~1
	Coefficients: 
		 [,1]     [,2]
	[1,]    0 10.71779
	Probalities at zero values of the covariates.
	2.214681e-05 0.9999779

	Transition model for state (component) 1 
	Model of type multinomial, formula: ~scale(Pacc)
	Coefficients: 
		 [,1]       [,2]
	[1,]    0 -0.9215182
	[2,]    0  1.8649734
	Probalities at zero values of the covariates.
	0.7153513 0.2846487

	Transition model for state (component) 2 
	Model of type multinomial, formula: ~scale(Pacc)
	Coefficients: 
		 [,1]     [,2]
	[1,]    0 2.471442
	[2,]    0 3.570856
	Probalities at zero values of the covariates.
	0.07788458 0.9221154

	Response model for response 1 
	Model of type gaussian, formula: rt ~ 1
	Coefficients: 
	[1] 5.512179
	sd  0.1921098

	Response model for response 1 
	Model of type gaussian, formula: rt ~ 1
	Coefficients: 
	[1] 6.3885
	sd  0.2402693

\end{verbatim}

The summary provides all parameters of the model, also the 
(redundant) zeroes for the base-line category in the multinomial model. 
The summary also prints the transition probabilities at the zero value 
of the covariate. Note that scaling of the covariate is useful in this 
regard as it makes interpretation of these intercept probabilities 
easier.

\subsection{Multivariate data}

Multivariate data can be modelled by providing a list of formulae as 
well as a list of family objects for the distributions of the various 
responses. In above examples we have only used the response times 
which were modelled with the gaussian distribution. The accuracy data 
are in the \code{speed} data are modelled with a multinomial by 
specifying the following: 
\begin{verbatim}
set.seed(1)
mod <- depmix(list(rt~1,corr~1), data=speed, nstates=2, 
	family=list(gaussian(),multinomial()), transition=~scale(Pacc), 
	instart=runif(2))
fm <- fit(mod)
\end{verbatim}
which provides the following fitted model parameters (only the 
response parameters are given here): 
\begin{verbatim}
	> summary(fm)
	\ldots 
	Response model(s) for state 1

	Response model for response 1 
	Model of type gaussian, formula: rt ~ 1
	Coefficients: 
	[1] 5.52169
	sd  0.2028857

	Response model for response 2 
	Model of type multinomial, formula: corr ~ 1
	Coefficients: 
		 [,1]      [,2]
	[1,]    0 0.1030554
	Probalities at zero values of the covariates.
	0.4742589 0.5257411

	Response model for response 1 
	Model of type gaussian, formula: rt ~ 1
	Coefficients: 
	[1] 6.39369
	sd  0.2373650

	Response model for response 2 
	Model of type multinomial, formula: corr ~ 1
	Coefficients: 
		 [,1]     [,2]
	[1,]    0 2.245514
	Probalities at zero values of the covariates.
	0.09573715 0.9042629 	
\end{verbatim}
As can be seen, state 1 has fast response times around chance level, 
whereas state 2 corresponds with slower responding at higher accuracy 
levels.

\subsection{Adding covariates on the prior probabilities}

The \pkg{depmixS4} contains another data set which is used to
illustrate the use of covariates on the prior probabilities of models,
in this case a latent class model. The \code{balance} data consists 
of 4 binary items (correct-incorrect) on a balance scale task 
\citet{Siegler198?}. The data form a subset of the data published in 
\citet{Jansen199?}.

Similarly to the transition matrix, covariates on the prior 
probabilities of the latent states (or classes in this case), are 
defined by using a one-sided formula: 
\begin{verbatim}
balance$age <- balance$age-5
set.seed(1)
mod <- mix(list(d1~1,d2~1,d3~1,d4~1), data=balance, nstates=2,
        family=list(multinomial(), multinomial(), multinomial(),
		multinomial()), respstart=c(rep(c(0.9,0.1),4),rep(c(0.1,0.9),4)), 
		prior=~age, initdata=balance)
fm <- fit(mod)
\end{verbatim}
Note here that we define a \code{mix} model instead of a \code{depmix}
models as these data form independent observations.

The summary of the \code{fit}ted model gives (only the prior model is 
shown here): 
\begin{verbatim}
> summary(fm)
Mixture probabilities model 
Model of type multinomial, formula: ~age
		[,1]   [,2]
[1,]    0 -2.5182573
[2,]    0  0.5512996
Probalities at zero values of the covariates.
0.9254119 0.07458815 
\end{verbatim}	
Hence at young ages, children have a high probability of incorrect 
responding in class~1, whereas the prior probability for class~2 
increases with age.

@@

Using package \pkg{Rdonlp2}, parameters may be fitted subject to
general linear (in-)equality constraints.  Constraining and fixing
parameters is done using the \code{conpat} argument to the
\code{depmix.fit}-function, which specifies for each parameter in the
model whether it's fixed (0) or free (1 or higher).  Equality
constraints can be imposed by having two parameters have the same
number in the \code{conpat} vector.  When only fixed values are
required the \code{fixed} argument can be used instead of
\code{conpat}, with zeroes for fixed parameters and other values (ones
e.g.) for non-fixed parameters.  Fitting the models subject to these
constraints is handled by the optimization routine \code{donlp2}.

\paragraph{Parameter numbering} When using the \code{conpat} and
\code{fixed} arguments, complete parameter vectors should be supplied,
i.e., these vectors should have length of the number of parameters of
the model, which can be obtained by calling \code{npar(object)}.
Parameters are numbered in the following order:
\begin{enumerate}
	\item  the prior model parameters
	\item  the parameters for the transition models
	\item  the response model parameters per state (and subsequently
	per response in the case of multivariate time series)
\end{enumerate}

To see the ordering of parameters use the following:
\begin{verbatim}
mod <- setpars(mod, value=1:npar(mod))
mod
\end{verbatim}

To see which parameters are fixed (by default only baseline parameters
in the multinomial logistic models for the transition models and the
initial state probabilities model):
\begin{verbatim}
mod <- setpars(mod, getpars(mod,which="fixed"))
mod
\end{verbatim}

When fitting constraints it is useful to have good starting values 
for the parameters and hence we first fit the following model: 
\begin{verbatim}
trst=c(0.9,0.1,0,0,0.1,0.9,0,0)
mod <- depmix(list(rt~1,corr~1),data=speed,transition=~Pacc,
	nstates=2,family=list(gaussian(),multinomial()),
	trstart=trst,inst=c(.999,0.001))
fm <- fit(mod)
\end{verbatim}
After this, we use the fitted values from this model to constrain the 
regression coefficients on the transition matrix:
\begin{verbatim}
# start with fixed and free parameters
conpat <- c(0,1,rep(c(0,1),4),1,1,0,1,1,1,0,1)
# constrain the beta's on the transition parameters to be equal
conpat[6] <- conpat[10] <- 2
fm <- fit(fm,equal=conpat)
\end{verbatim}
Using \code{summary} on the fitted model shows that the regression 
coefficients are now estimated at the same value of 12.66. 
Note that these arguments provide the possibility for arbitrary 
constraints, also between, e.g., a multinomial regression coefficient 
for the transition matrix and the mean of a gaussian response model. 
Whether such constraints make sense is hence the responsibility of 
the user.

@@

The \pkg{depmixS4} package was built with the aim of having the 
possibility of adding new response distributions (and possibly also 
other models for the prior and transition probabilities models). 
Therefore, the EM algorithm simply calls \code{fit} functions from the 
separate response models without knowing what they are. As a 
consequence, adding user-specified response models is straightforward. 
User-defined distributions should extend the \code{response}-class 
and have the following slots:
\begin{enumerate}
	\item y: the response variable
	\item x: the design matrix, possibly only an intercept
	\item paramaters: a named list with the coefficients and possibly 
	other parameters, e.g., the standard deviation in the gaussian 
	response model
	\item fixed: a vector of logicals indicating whether parameters are 
	fixed
	\item npar: numerical indicating the number of parameters of the model
\end{enumerate}

In the \code{speed} data example, it may be more appropriate to model
the response times as an exgaus distribution rather than using the
gaussian.  We can do so as follows, first we define our own
exgaus-class extending the response-class and set it to be a generic:
\begin{verbatim}
	setClass("exgaus", contains="response")
	setGeneric("exgaus", function(y, pstart = NULL, fixed = NULL, ...) 
	standardGeneric("exgaus"))
\end{verbatim}

The so-defined class now needs a number of methods: 
\begin{enumerate}
	\item constructor: function to create instances of the class 
	with starting values
	\item show method: to print the model to the terminal
	\item dens: the function that computes the density of the responses
	\item getpars and setpars: to get and set parameters 
	\item predict: to generate predict'ed values 
	\item fit: function to fit the model using posterior weights (used 
	by the EM algorithm)
\end{enumerate}

Only the constructor function and the fit function are provided here
and the complete code is provided in the helpfile for \code{makeDepmix}. 
The \code{exgaus} example uses the \pkg{gamlss.distr} package for 
computing the \code{dens}ity and for fitting the parameters.

The constructor function is defined as: 
\begin{verbatim}
setMethod("exgaus", 
	signature(y="ANY"), 
	function(y,pstart=NULL,fixed=NULL, ...) {
		y <- matrix(y,length(y))
		x <- matrix(1)
		parameters <- list()
		npar <- 3
		if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
		if(!is.null(pstart)) {
			if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
			parameters$mu <- pstart[1]
			parameters$sigma <- log(pstart[2])
			parameters$nu <- log(pstart[3])
		}
		mod <- new("exgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar)
		mod
	}
)
\end{verbatim}

The \code{fit} function is defined as follows: 
\begin{verbatim}
setMethod("fit","exgaus",
	function(object,w) {
		if(missing(w)) w <- NULL
		y <- object@y
		fit <- gamlss(y~1,weights=w,family=exGAUS(),
			control=gamlss.control(n.cyc=100,trace=FALSE),
			mu.start=object@parameters$mu,
			sigma.start=exp(object@parameters$sigma),
			nu.start=exp(object@parameters$nu))
		pars <- c(fit$mu.coefficients,fit$sigma.coefficients,fit$nu.coefficients)
		object <- setpars(object,pars)
		object
	}
)
The \code{fit} function defines a trivial \code{gamlss} model with 
only an intercept to be estimated and then sets the fitted parameters 
back into their respective slots in the `exgaus' object. See the help 
for \pkg{gamlss.distr} for interpretation of these parameters.

After defining all the necessary methods for the new response model, 
we can  now define the dependent mixture model using this reponse. 
The function \code{makeDepmix} is added to \pkg{depmixS4} to have 
full control over model specification, and we need it here.

We first create all the response models that we want as a double list: 
\begin{verbatim}
rModels <- list(
	list(
		exgaus(rt,pstart=c(5,.1,.1)),
		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(0.5,0.5))
	),
	list(
		exgaus(rt,pstart=c(6,.1,.1)),
		GLMresponse(formula=corr~1,data=speed,family=multinomial(),pstart=c(.1,.9))
	)
)
\end{verbatim}
Next, we define the transition and prior probability models using the 
transInit function (which produces a transInit model, which also extends 
the response class): 
\begin{verbatim}
trstart=c(0.9,0.1,0.1,0.9)
transition <- list()
transition[[1]] <- transInit(~Pacc,nstates=2,data=speed,pstart=c(0.9,0.1,0,0))
transition[[2]] <- transInit(~Pacc,nstates=2,data=speed,pstart=c(0.1,0.9,0,0))
inMod <- transInit(~1,ns=2,pstart=c(0.1,0.9),data=data.frame(1))
\end{verbatim}
Finally, we put everything together using \code{makeDepmix} and fit 
the model: 
\begin{verbatim}
mod <- makeDepmix(response=rModels,transition=transition,
prior=inMod,stat=FALSE)
fm <- fit(mod)
\end{verbatim}

Using \code{summary} will print the fitted parameters. Note that the 
use of \code{makeDepmix} allows the possibility of, say, fitting a 
gaussian in one state and an exgaus distribution in another state.
